Reduced plasticity and opportunity for selection in terrestrial ectotherm populations under climate change

### 2023-12-04 ms.qmd
Updated MS

Climate change is expected to result in warmer and more variable thermal environments for many organisms globally [@Suarez2023; @Easterling2000; @Ummenhofer2017]. Greater thermal variability in the past is expected to have resulted in strong selection pressures leading to genetic adaptation and/or the evolution of adaptive phenotypic plasticity -- both of which are considered important for population resilience to human-induced climate change [@Seebacher2015-wi; @Merila2014; @Chevin2015; @Chevin2017; @Chevin2010; @Nunney2016]. The extent to which adaptive plasticity and genetic adaptation will allow organisms to adapt to new environmental conditions is not yet clearly established [@Chevin2017]. Without such responses high rates of extinction are predicted if organisms cannot track suitable habitats [@Cahill2012; @Nunney2016]. Phenotypic plasticity is expected to the the 'first line of defence' against changing climates by buyding time for genetic adaptation to take place (i.e., the 'plasticty first hypothesis' - ). Phenotypic plasticity is predicted to evolve when environmental variability is high but predictable and the costs of plasticity are low [@Chevin2017; @Scheiner2020; @Nunney2016]. Despite this theoretical expectation, empirical support is scant [], likely because many organisms can behaviorally adjust micro-habitat selection to offset thermal stress, the costs of plasticity are high [@Chevin2015], and/or the prediction is likely only supported for specific plastic responses (i.e., active and developmental plasticity). Reversible forms of phenotypic plasticity, such as acclimatization, may not be expected to adhere to such predictions.

Arguably, reversible forms of phenotypic plasticty, such as acclimitisation and behavioural plasticity, are likely to provide greater potential to buffer populations from climate impacts as responses can be fine-tuned to environments [assuming the costs of plasticty are low – @Scheiner2020]. Such responses are driven by changes in underlying physiology that can respond rapidly to the environment. However, studies have primarily been focused on mean physiological responses, paying little attention to changes in physiological variability [@Seebacher2015-wi; @Havird2020-mj]. For example, most work has focused on the degree to which mean thermal tolerances or acclimation abilities are likely to shift in response to thermal environments [@Seebacher2015-wi; @Gunderson2015-hz; @Havird2020-mj; @Pottier2022], neglecting how intrapopulation variability might also be impacted. Understanding how variability in physiological rates -- traits thought to be closely closely linked to fitness -- are affected by climate change has important implications for understanding a populations capacity for physiological trait evolution because lack of physiological variation is expected to play a role in limiting responses to selection (i.e., the 'opportunity for selection on a trait') [@Pelletier2012]. Change in physiological variability can also inform on the extent to which populations are able to buffer against environmental change [for example, through 'portfolio effects' see @Schindler2010]. Decreases in phenotypic variance may also suggest strong stabilising selection or constraints on performance depedning on the trait [@Scheiner2020]. Changes in phenotypic variability can also result in 'phenotypic drift' that may result in the loss of important variation relevant for adaptation.

@@

Here, we use meta-analysis and new effect sizes to re-evaluate the degree to which aquatic and terrestrial ectotherms are capable of physiological plasticity. We then expand these effect sizes to allow for comparisons of variance in physiological rates to ask the following questions: 1) How much is variance in physiological rates expected to change, if at all, as temperatures rise?; 2) Are temperture effects on plastic adjustments in physiolgical rates larger than changes in variance across aquatic and terrestrial ectotherms?; 3) Are changes in plasticity or variance in physiological rates impacted by a population's past climate history? and 4) How are means and variances in physiological rates expected to change under climate change?

>>>>

Climate change is expected to result in warmer and more variable thermal environments for many organisms globally [@Suarez2023; @Easterling2000; @Ummenhofer2017]. Greater thermal variability in the past is expected to have resulted in strong selection pressures leading to genetic adaptation and/or the evolution of adaptive phenotypic plasticity -- both of which are considered important for population resilience to contemporary human-induced climate change [@Seebacher2015-wi; @Merila2014; @Chevin2015; @Chevin2017; @Chevin2010; @Nunney2016]. The extent to which adaptive plasticity and genetic adaptation will allow organisms to adapt to new environmental conditions is not yet clearly established [@Chevin2017]. Without such responses, high rates of extinction are predicted if organisms cannot track suitable habitats [@Cahill2012; @Nunney2016]. Phenotypic plasticity is expected to the the 'first line of defence' against changing climates by buying time for genetic adaptation to take place (i.e., the 'plasticty first hypothesis' - ). Phenotypic plasticity is predicted to evolve when environmental variability is high but predictable and the costs of plasticity is low [@Chevin2017; @Scheiner2020; @Nunney2016]. Despite this theoretical expectation, empirical support is scant [], likely because many organisms can behaviorally adjust micro-habitat selection to offset thermal stress, the costs of plasticity are high [@Chevin2015], and/or the prediction is likely only supported for specific plastic responses (i.e., active and developmental plasticity). Reversible forms of phenotypic plasticity, such as acclimatization, may not be expected to adhere to such predictions.

Arguably, reversible forms of phenotypic plasticty, such as acclimitisation and behavioural plasticity, are likely to provide greater potential to buffer populations from climate impacts as responses can be fine-tuned to environments [assuming the costs of plasticty are low – @Scheiner2020]. Such responses are driven by changes in underlying physiology that can respond rapidly to the environment. However, studies have primarily been focused on mean physiological responses, paying little attention to changes in physiological variability [@Seebacher2015-wi; @Havird2020-mj]. For example, most work has focused on the degree to which mean thermal tolerances or acclimation abilities are likely to shift in response to thermal environments [@Seebacher2015-wi; @Gunderson2015-hz; @Havird2020-mj; @Pottier2022], neglecting how intrapopulation variability might also be impacted. Understanding how variability in physiological rates -- traits thought to be closely closely linked to fitness -- are affected by climate change has important implications for understanding a populations capacity for physiological trait evolution because lack of physiological variation is expected to play a role in limiting responses to selection (i.e., the 'opportunity for selection on a trait') [@Pelletier2012]. Change in physiological variability can also inform on the extent to which populations are able to buffer against environmental change [for example, through 'portfolio effects' see @Schindler2010]. Decreases in phenotypic variance may also suggest strong stabilising selection or constraints on performance depending on the trait [@Scheiner2020]. The implications of changes in variance could therefore have wide-reaching implications for understanding the capacity for populations to adapt to novel environments [@Scheiner2020].

@@

Here, we use meta-analysis and new effect sizes to re-evaluate the degree to which aquatic and terrestrial ectotherms are capable of physiological plasticity. We then expand these effect sizes to allow for comparisons of variance in physiological rates to ask the following questions: 1) How much is variance in physiological rates expected to change, if at all, as temperatures rise? 2) Are temperture effects on plastic adjustments in physiolgical rates larger than changes in variance across aquatic and terrestrial ectotherms? 3) Are changes in plasticity or variance in physiological rates impacted by a population's past climate history? and 4) How are means and variances in physiological rates expected to change under climate change?
